Free Theorems for Nested Types

### 2020-06-23 free-theorems-nesteds/paper3.tex
changes to paper3

\begin{dfn}\label{def:set-sem}
%Let $\rho$ be a set environment.
The {\em set interpretation} $\setsem{\cdot} : \F \to [\setenv, \set]$
is defined by
\begin{align*}
  \setsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
  \setsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
%  \setsem{\Gamma;\emptyset \vdash v} \rho &= \rho v \mbox{
%    if } v \in \tvars^0 \\ 
  \setsem{\Gamma; \ol\gamma
    %\emptyset
    \vdash \Nat^{\ol{\alpha}}
    \,F\,G}\rho &= \{\eta : \lambda \ol{A}. \,\setsem{\Gamma;
    \ol{\alpha} \vdash
    F}\rho[\ol{\alpha := A}] 
      \Rightarrow \lambda \ol{A}.\,\setsem{\Gamma; \ol\gamma,
        \ol{\alpha} \vdash G}\rho[\ol{\alpha := A}] \\ 
      &\hspace{0.3in}|~\forall \overline{A}, \overline{B} :
      \set. \forall \overline{R : \rel(A, B)}.\\ 
      &\hspace{0.4in}(\eta_{\overline{A}}, \eta_{\overline{B}})
      : \relsem{\Gamma; \ol{\alpha} \vdash F}\Eq_{\rho}[\ol{\alpha := R}]
      \rightarrow \relsem{\Gamma;\ol\gamma, \ol{\alpha} \vdash
        G}\Eq_{\rho}[\ol{\alpha := R}] \} \\
  \setsem{\Gamma;\Phi \vdash \phi\ol{F}}\rho &=
  (\rho\phi)\,\ol{\setsem{\Gamma;\Phi \vdash
    F}\rho}\\
  \setsem{\Gamma;\Phi \vdash F+G}\rho &=
  \setsem{\Gamma;\Phi \vdash F}\rho +
  \setsem{\Gamma;\Phi \vdash G}\rho\\
  \setsem{\Gamma;\Phi \vdash F\times G}\rho &=
  \setsem{\Gamma;\Phi \vdash F}\rho \times
  \setsem{\Gamma;\Phi \vdash G}\rho\\ 
  \setsem{\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
    \ol{\alpha}. H)\ol{G}}\rho &= (\mu
    T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho}\\
    \text{where } T^\set_{H,\rho}\,F & = \lambda
  \ol{A}. \setsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
    H}\rho[\phi :=  F][\ol{\alpha := A}]\\
  \text{and } T^\set_{H,\rho}\,\eta &= \lambda
  \ol{A}. \setsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
    H}\id_\rho[\phi := \eta][\ol{\alpha := \id_{A}}]
\end{align*}
\end{dfn}
The interpretations in Definition~\ref{def:set-sem} respect weakening,
i.e., a type and its weakenings have the same set interpretations.
The same holds for the actions of these interpretations on morphisms
in Definition~\ref{def:set-sem-funcs} below.
%= \setsem{\Gamma \vdash \sigma} \rho \to
%\setsem{\Gamma \vdash \tau} \rho$, as expected {\color{blue} by IEL}.
If $\rho$ is a set environment and $\vdash F$ then we may
write $\setsem{\vdash F}$ instead of $\setsem{\vdash F}\rho$
since the environment is immaterial.
We note that the third clause of
Definition~\ref{def:set-sem} does indeed define a set: local finite
presentability of $\set$ and $\omega$-cocontinuity of
$\setsem{\Gamma;\ol{\alpha} \vdash F}\rho$ ensure that $\{\eta :
\setsem{\Gamma;\ol{\alpha} \vdash F}\rho \Rightarrow
\setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho\}$ (which contains
$\setsem{\Gamma;\ol\gamma \vdash \Nat^{\ol{\alpha}}\,F\,G}\rho$) is a
subset of $\big\{({\setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho[\ol{\alpha
      := S}]})^{(\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[\ol{\alpha
      := S}])}~\big|~ \ol{S} = (S_1,...,S_{|\ol{\alpha}|}), \mbox{ and
}$ $S_i \mbox{ is a finite set for } i =
1,...,|\ol{\alpha}|\big\}$. There are countably many choices for
tuples $\ol{S}$, and each of these gives rise to a morphism from
${\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[ \ol{\alpha := S}]}$ to
${\setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho[\ol{\alpha := S}]}$. But
there are only $\set$-many choices of morphisms between any
%these (or any)
two objects since $\set$ is locally small. Interpretations of
$\Nat$ types in Definition~\ref{def:set-sem} ensure $\setsem{\Gamma
  \vdash F \to G}$ and $\setsem{\Gamma \vdash \forall
  \alpha. F}$ are as expected in any parametric model.

In order to make sense of the last clause in
Definition~\ref{def:set-sem}, we need to know that, for each $\rho \in
\setenv$, $T^\set_{H,\rho}$ is an $\omega$-cocontinuous endofunctor on
$[\set^k, \set]$, and thus admits a fixpoint.  Since
$T_{H,\rho}^\set$ is defined in terms of $\setsem{\Gamma;\ol\gamma,\phi,
  \ol{\alpha} \vdash H}$, this means that interpretations of types
must be such functors, which in turn means that the actions of set
interpretations of types on objects and on morphisms in $\setenv$ are
intertwined. Fortunately, we know from~\cite{jp19} that, for every
$\Gamma; \ol{\alpha} \vdash G$, $\setsem{\Gamma; \ol{\alpha} \vdash
  G}$ is actually in $[\set^k,\set]$ where $k = |\ol \alpha|$. This
means that for each $\setsem{\Gamma; \ol\gamma, \phi^k, \ol{\alpha} \vdash
  H}$, the corresponding operator $T^\set_{H}$ can be extended to a
{\em functor} from $\setenv$ to $[[\set^k,\set],[\set^k,\set]]$. The
action of $T^\set_H$ on an object $\rho \in \setenv$ is given by the
higher-order functor $T_{H,\rho}^\set$, whose actions on objects
(functors in $[\set^k, \set]$) and morphisms (natural transformations
between such functors) are given in Definition~\ref{def:set-sem}. Its
action on a morphism $f : \rho \to \rho'$ is the higher-order natural
transformation $T^\set_{H,f} : T^\set_{H,\rho} \to T^\set_{H,\rho'}$
whose action on $F : [\set^k,\set]$ is the natural transformation
$T^\set_{H,f}\, F : T^\set_{H,\rho}\,F \to T^\set_{H,\rho'}\,F$ whose
component at $\ol{A}$ is $(T^\set_{H,f}\, F)_{\ol{A}} =
\setsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi :=
  \id_F][\ol{\alpha := \id_A}]$. The next definition uses the functor
$T^\set_H$ to define the actions of functors interpreting types on
morphisms between set environments.

@@

@@

The interpretations in Definition~\ref{def:rel-sem}, as well as in
Definition~\ref{def:rel-sem-funcs} below, respect weakening, and also
ensure that $\relsem{\Gamma \vdash F \to G}$ and $\relsem{\Gamma
  \vdash \forall \alpha. F}$ are as expected in any parametric model.
%$\relsem{\Gamma \vdash
%  \sigma \to \tau}\rho = \relsem{\Gamma \vdash \sigma} \rho \to
%\relsem{\Gamma \vdash \tau} \rho$.
If $\rho$ is a relational environment and $\vdash F$, then we write
$\relsem{\vdash F}$ instead of $\relsem{\vdash F}\rho$ as for set
interpretations.  For the last clause in Definition~\ref{def:rel-sem}
to be well-defined we need $T_{H,\rho}$ to be an $\omega$-cocontinuous
endofunctor on $RT$ so that, by Lemma~\ref{lem:fp}, it admits a
fixpoint. Since $T_{H,\rho}$ is defined in terms of
$\relsem{\Gamma;\ol\gamma,\phi^k, \ol{\alpha} \vdash H}$, this means that
relational interpretations of types must be $\omega$-cocontinuous
functors from $\relenv$ to $RT_0$, which in turn entails that the
actions of relational interpretations of types on objects and on
morphisms in $\relenv$ are intertwined. As for set interpretations, we
know from~\cite{jp19} that, for every $\Gamma; \ol{\alpha} \vdash F$,
$\setsem{\Gamma; \ol{\alpha} \vdash F}$ is actually in
$[\rel^k,\rel]$ where $k = |\ol \alpha|$.
%In fact, we already know from~\cite{jp19} that, for every $\Gamma;
%\ol{\alpha} \vdash \tau : \F$, $\relsem{\Gamma; \ol{\alpha} \vdash
%  \tau}$ is actually functorial in $\ol{\alpha}$ and
%$\omega$-cocontinuous.
We first define the actions of each of these functors on morphisms
between environments in Definition~\ref{def:rel-sem-funcs}, and then
argue that the functors given by Definitions~\ref{def:rel-sem}
and~\ref{def:rel-sem-funcs} are well-defined and have the required
properties. To do this, we extend $T_H$ to a {\em functor} from
$\relenv$ to $[[\rel^k,\rel],[\rel^k,\rel]]$. Its action on an object
$\rho \in \relenv$ is given by the higher-order functor $T_{H,\rho}$
whose actions on objects and morphisms are given in
Definition~\ref{def:rel-sem-funcs}. Its action on a morphism $f : \rho
\to \rho'$ is the higher-order natural transformation $T_{H,f} :
T_{H,\rho} \to T_{H,\rho'}$ whose action on any $F : [\rel^k,\rel]$ is
the natural transformation $T_{H,f}\, F : T_{H,\rho}\, F \to
T_{H,\rho'}\, F$ whose component at $\ol{R}$ is $(T_{H,f}\,
F)_{\ol{R}} = \relsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi :=
  \id_F][\ol{\alpha := \id_R}]$.  The next definition uses the functor
$T_H$ to define the actions of functors interpreting types on
morphisms between relation environments.

@@

Define, for $\Gamma; \ol{\alpha} \vdash F$, the term {\color{blue}
  $\id_F$} to be $\Gamma;\emptyset~|~\emptyset \vdash
L_{\ol{\alpha}}x.x : \Nat^{\ol{\alpha}} F\,F$ and, for terms $\Gamma;
\emptyset \,|\, \Delta \vdash t: \Nat^{\overline{\alpha}} F\,G$ and
$\Gamma; \emptyset \,|\, \Delta \vdash s: \Nat^{\overline{\alpha}}
G\,H$, the {\em composition} $s \circ t$ of $t$ and $s$ to be $\Gamma;
\emptyset \,|\, \Delta \vdash L_{\overline{\alpha}}
x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
\Nat^{\overline{\alpha}} F\,H$. Then $\setsem{\Gamma; \emptyset \,|\,
  \emptyset \vdash \id_{F} : \Nat^{\ol{\alpha}} F\,F} \rho\, \ast =
\id_{\lambda \ol{A}. \setsem{\Gamma; \ol{\alpha} \vdash F} \rho
  [\ol{\alpha := A}]}$ for any set environment $\rho$, and
$\setsem{\Gamma; \emptyset \,|\, \Delta \vdash s \circ t:
  \Nat^{\overline{\alpha}} F\,H} = \setsem{\Gamma; \emptyset \,|\,
  \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H} \circ
\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t:
  \Nat^{\overline{\alpha}} F\,G}$.  Straightforward calculation using
these equalities shows that terms of $\Nat$ type behave as

@@

@@

\hspace*{0.135in}\setsem{\Gamma; {\color{blue} \ol\alpha} \,|\, x :
  \Nat^{\overline{\alpha}, 
    \overline{\gamma}} F\,G, \overline{y : \Nat^{\overline{\gamma}}
    K\, H} \vdash (\map_G^{\overline{K},
    \overline{H}} \,\overline{y}) \circ
  (L_{\overline{\gamma}} z. x_{\overline{K}, \overline{\gamma}}
  z) : \Nat^{\overline{\gamma}} F[\overline{\alpha := K}]\,

@@

= \setsem{ \Gamma; {\color{blue} \ol\alpha} \,|\, x
  : \Nat^{\overline{\alpha}, \overline{\gamma}} F\, G, \overline{y :
    \Nat^{\overline{\gamma}} K\, H} \vdash
  (L_{\overline{\gamma}} z. x_{\overline{H}, \overline{\gamma}} z)
  \circ (\map_F^{\overline{K}, \overline{\tau}} \,
  \overline{y}) : \Nat^{\overline{\gamma}} F[\overline{\alpha :=

@@

When $x = \mathit{in}_H$ and $\xi = \Nat^{\overline{\gamma}} H[\phi :=

@@

\overline{\alpha}. H)\overline{G}$ Theorem~\ref{eq:ft-from-nat}

@@

\hspace*{0.135in}\setsem{\Gamma; {\color{blue} \ol\alpha} \,|\, \overline{y :
    \Nat^{\overline{\gamma}} 
    F\, G} \vdash (\map_{(\mu \phi. \lambda

@@

    \overline{G}}\, \overline{y}) \circ
  (L_{\overline{\gamma}} z. (\tin_{H})_{\overline{F},

@@

= \setsem{\Gamma; {\color{blue} \ol\alpha} \,|\, \overline{y :
  \Nat^{\overline{\gamma}} F\, G} 
  \vdash (L_{\overline{\gamma}} z. (\tin_H)_{\overline{G}, \overline{\gamma}} z)
  \circ (\map_{H[\phi := (\mu \phi. \lambda \overline{\alpha}. H)

@@

    \overline{G}}\, \overline{y}) 

@@

  &\setsem{\emptyset; \alpha~|~ y : \Nat^\emptyset F \, G
    \vdash (\map_{\mathit{Bush}\,\alpha}^{{F},
      {G}}\, y) \circ (L_{{\emptyset}}
    z. \mathit{reverseBush}_{F}\, z) : \Nat^{{\emptyset}}

@@

  &\setsem{\emptyset; \alpha~|~ y : \Nat^\emptyset F\, G
    \vdash (L_{\emptyset} z. \mathit{reverseBush}_{G} \,z) \circ
    (\map_{\mathit{Bush} \, \alpha}^{F, G}\,
           {y}) : \Nat^{\emptyset} (\mathit{Bush} \, F ) \,

@@

  &\setsem{\emptyset; \alpha,\gamma~|~y : \Nat^\emptyset F \,
    G \vdash (\map_{\mathit{Tree}\,\alpha \, \gamma}^{F,
      G}\, y) \circ (L_{{\gamma}}
    z. \mathit{reverseTree}_{F, \gamma}\, z) : \Nat^{{\gamma}}

@@

    \gamma) }\\ =\; &\setsem{\emptyset; \alpha,\gamma~|~y :
    \Nat^\emptyset F \, G \vdash (L_{\gamma}
    z. \mathit{reverseTree}_{G, \gamma} \,z) \circ
    (\map_{\mathit{Tree} \, \alpha\, \gamma}^{{F},
      {G}}\, y) : \Nat^{\gamma} (\mathit{Tree} \,

@@

We also have the expected relationships between the interpetations of
$\map$, $\tin$, and $\fold$, namely

@@

\vspace*{-0.3in}

@@

\[\hspace*{-3.6in} = \setsem{\Gamma; \emptyset \,|\, \emptyset \vdash \Id_{(\mu
    \phi. \lambda \overline{\alpha}. H)\overline{\beta}} : \xi}\]

@@

\[\hspace*{-3.32in} =\setsem{ \Gamma; \emptyset \,|\, \emptyset \vdash
  \Id_{H[\phi := (\mu \phi. \lambda
      \overline{\alpha}. H)\overline{\beta}]} : \xi}\]

@@

>>>>

\begin{dfn}\label{def:set-sem}
%Let $\rho$ be a set environment.
The {\em set interpretation} $\setsem{\cdot} : \F \to [\setenv, \set]$
is defined by
\begin{align*}
  \setsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
  \setsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
%  \setsem{\Gamma;\emptyset \vdash v} \rho &= \rho v \mbox{
%    if } v \in \tvars^0 \\ 
  \setsem{\Gamma; \ol\gamma
    %\emptyset
    \vdash \Nat^{\ol{\alpha}}
    \,F\,G}\rho &= \{\eta : \lambda \ol{A}. \,\setsem{\Gamma;
    \ol{\alpha} \vdash
    F}\rho[\ol{\alpha := A}] 
      \Rightarrow \lambda \ol{A}.\,\setsem{\Gamma; \ol\gamma,
        \ol{\alpha} \vdash G}\rho[\ol{\alpha := A}] \\ 
      &\hspace{0.3in}|~\forall \overline{A}, \overline{B} :
      \set. \forall \overline{R : \rel(A, B)}.\\ 
      &\hspace{0.4in}(\eta_{\overline{A}}, \eta_{\overline{B}})
      : \relsem{\Gamma; \ol{\alpha} \vdash F}\Eq_{\rho}[\ol{\alpha := R}]
      \rightarrow \relsem{\Gamma;\ol\gamma, \ol{\alpha} \vdash
        G}\Eq_{\rho}[\ol{\alpha := R}] \} \\
  \setsem{\Gamma;\Phi \vdash \phi\ol{F}}\rho &=
  (\rho\phi)\,\ol{\setsem{\Gamma;\Phi \vdash
    F}\rho}\\
  \setsem{\Gamma;\Phi \vdash F+G}\rho &=
  \setsem{\Gamma;\Phi \vdash F}\rho +
  \setsem{\Gamma;\Phi \vdash G}\rho\\
  \setsem{\Gamma;\Phi \vdash F\times G}\rho &=
  \setsem{\Gamma;\Phi \vdash F}\rho \times
  \setsem{\Gamma;\Phi \vdash G}\rho\\ 
  \setsem{\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
    \ol{\alpha}. H)\ol{G}}\rho &= (\mu
    T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho}\\
    \text{where } T^\set_{H,\rho}\,F & = \lambda
  \ol{A}. \setsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
    H}\rho[\phi :=  F][\ol{\alpha := A}]\\
  \text{and } T^\set_{H,\rho}\,\eta &= \lambda
  \ol{A}. \setsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
    H}\id_\rho[\phi := \eta][\ol{\alpha := \id_{A}}]
\end{align*}
\end{dfn}
%The interpretations in Definition~\ref{def:set-sem} respect weakening,
%i.e., a type and its weakenings have the same set interpretations.
%
%= \setsem{\Gamma \vdash \sigma} \rho \to
%\setsem{\Gamma \vdash \tau} \rho$, as expected {\color{blue} by IEL}.
If $\rho \in \setenv$ and $\vdash F$ then we write $\setsem{\vdash F}$
instead of $\setsem{\vdash F}\rho$ since the environment is
immaterial. The third clause of Definition~\ref{def:set-sem} does
indeed define a set: local finite presentability of $\set$ and
$\omega$-cocontinuity of $\setsem{\Gamma;\ol{\alpha} \vdash F}\rho$
ensure that $\{\eta : \setsem{\Gamma;\ol{\alpha} \vdash F}\rho
\Rightarrow \setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho\}$
(which contains $\setsem{\Gamma;\ol\gamma \vdash
  \Nat^{\ol{\alpha}}\,F\,G}\rho$) is a subset of
$\big\{({\setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho[\ol{\alpha
      := S}]})^{(\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[\ol{\alpha
      := S}])}~\big|~ \ol{S} = (S_1,...,S_{|\ol{\alpha}|}), \mbox{ and
}$ $S_i \mbox{ is a finite set for } i =
1,...,|\ol{\alpha}|\big\}$. There are countably many choices of tuples
$\ol{S}$, each of which gives rise to a morphism from
${\setsem{\Gamma;\ol{\alpha} \vdash F}\rho[ \ol{\alpha := S}]}$ to
${\setsem{\Gamma;\ol\gamma,\ol{\alpha} \vdash G}\rho[\ol{\alpha :=
      S}]}$. But there are only $\set$-many choices of morphisms
between any two objects since $\set$ is locally small.  That
$\setsem{\Gamma; \ol\gamma \vdash \Nat^\alpha F\,G}$ is
$\omega$-cocontinuous follows from $\omega$-cocontinuity of
$\setsem{\Gamma;\ol\alpha \vdash F}$ and
$\setsem{\Gamma;\ol\gamma,\ol\alpha \vdash G}$. Interpretations of
$\Nat$-types ensure that $\setsem{\Gamma \vdash F \to G}$ and
$\setsem{\Gamma \vdash \forall \alpha. F}$ are as expected in any
parametric model.

To make sense of the last clause in
Definition~\ref{def:set-sem}, we need to know that, for each $\rho \in
\setenv$, $T^\set_{H,\rho}$ is an $\omega$-cocontinuous endofunctor on
$[\set^k, \set]$, and thus admits a fixpoint.  Since
$T_{H,\rho}^\set$ is defined in terms of $\setsem{\Gamma;\ol\gamma,\phi,
  \ol{\alpha} \vdash H}$, this means that interpretations of types
must be such functors, which in turn means that the actions of set
interpretations of types on objects and on morphisms in $\setenv$ are
intertwined. Fortunately, we know from~\cite{jp19} that, for every
$\Gamma; \ol{\alpha} \vdash G$, $\setsem{\Gamma; \ol{\alpha} \vdash
  G}$ is actually in $[\set^k,\set]$ where $k = |\ol \alpha|$. This
means that for each $\setsem{\Gamma; \ol\gamma, \phi^k, \ol{\alpha} \vdash
  H}$, the corresponding operator $T^\set_{H}$ can be extended to a
{\em functor} from $\setenv$ to $[[\set^k,\set],[\set^k,\set]]$. The
action of $T^\set_H$ on an object $\rho \in \setenv$ is given by the
higher-order functor $T_{H,\rho}^\set$, whose actions on objects
(functors in $[\set^k, \set]$) and morphisms (natural transformations
between such functors) are given in Definition~\ref{def:set-sem}. Its
action on a morphism $f : \rho \to \rho'$ is the higher-order natural
transformation $T^\set_{H,f} : T^\set_{H,\rho} \to T^\set_{H,\rho'}$
whose action on $F : [\set^k,\set]$ is the natural transformation
$T^\set_{H,f}\, F : T^\set_{H,\rho}\,F \to T^\set_{H,\rho'}\,F$ whose
component at $\ol{A}$ is $(T^\set_{H,f}\, F)_{\ol{A}} =
\setsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi :=
  \id_F][\ol{\alpha := \id_A}]$. The next definition uses the functor
$T^\set_H$ to define the actions of functors interpreting types on
morphisms between set environments.

\begin{comment}
To make sense of the last clause in
Definition~\ref{def:set-sem}, we need to know that, for each $\rho \in
\setenv$, $T^\set_{H,\rho}$ is an $\omega$-cocontinuous endofunctor on
$[\set^k, \set]$, and thus admits a fixpoint.  Since
$T_{H,\rho}^\set$ is defined in terms of $\setsem{\Gamma;\ol\gamma,\phi,
  \ol{\alpha} \vdash H}$, this means that interpretations of types
must be such functors, which in turn means that the actions of set
interpretations of types on objects and on morphisms in $\setenv$ are
intertwined. Fortunately, we know from~\cite{jp19} that, for every
$\Gamma; \ol{\alpha} \vdash G$, $\setsem{\Gamma; \ol{\alpha} \vdash
  G}$ is actually in $[\set^k,\set]$ where $k = |\ol \alpha|$. This
means that for each $\setsem{\Gamma; \ol\gamma, \phi^k, \ol{\alpha} \vdash
  H}$, the corresponding operator $T^\set_{H}$ can be extended to a
{\em functor} from $\setenv$ to $[[\set^k,\set],[\set^k,\set]]$. The
action of $T^\set_H$ on an object $\rho \in \setenv$ is given by the
higher-order functor $T_{H,\rho}^\set$, whose actions on objects
(functors in $[\set^k, \set]$) and morphisms (natural transformations
between such functors) are given in Definition~\ref{def:set-sem}. Its
action on a morphism $f : \rho \to \rho'$ is the higher-order natural
transformation $T^\set_{H,f} : T^\set_{H,\rho} \to T^\set_{H,\rho'}$
whose action on $F : [\set^k,\set]$ is the natural transformation
$T^\set_{H,f}\, F : T^\set_{H,\rho}\,F \to T^\set_{H,\rho'}\,F$ whose
component at $\ol{A}$ is $(T^\set_{H,f}\, F)_{\ol{A}} =
\setsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash H}f[\phi :=
  \id_F][\ol{\alpha := \id_A}]$. The next definition uses the functor
$T^\set_H$ to define the actions of functors interpreting types on
morphisms between set environments.
\end{comment}

@@

Definitions~\ref{def:set-sem} and~\ref{def:set-sem-funcs} respect
weakening, i.e., ensure that a type and its weakenings have the same
set interpretations.

@@

The interpretations in Definitions~\ref{def:rel-sem}
and~\ref{def:rel-sem-funcs} below respect weakening, and also ensure
that $\relsem{\Gamma \vdash F \to G}$ and $\relsem{\Gamma \vdash
  \forall \alpha. F}$ are as expected in any parametric model.
%$\relsem{\Gamma \vdash
%  \sigma \to \tau}\rho = \relsem{\Gamma \vdash \sigma} \rho \to
%\relsem{\Gamma \vdash \tau} \rho$.
That $\relsem{\Gamma; \ol\gamma \vdash \Nat^\alpha F\,G}$ is
$\omega$-cocontinuous follows from $\omega$-cocontinuity of
$\relsem{\Gamma;\ol\alpha \vdash F}$ and
$\relsem{\Gamma;\ol\gamma,\ol\alpha \vdash G}$.  If $\rho \in \relenv$
and $\vdash F$, then we write $\relsem{\vdash F}$ instead of
$\relsem{\vdash F}\rho$.  For the last clause in
Definition~\ref{def:rel-sem} to be well-defined we need $T_{H,\rho}$
to be an $\omega$-cocontinuous endofunctor on $RT$ so that, by
Lemma~\ref{lem:fp}, it admits a fixpoint. Since $T_{H,\rho}$ is
defined in terms of $\relsem{\Gamma;\ol\gamma,\phi^k, \ol{\alpha}
  \vdash H}$, this means that relational interpretations of types must
be $\omega$-cocontinuous functors from $\relenv$ to $RT_0$, which in
turn entails that the actions of relational interpretations of types
on objects and on morphisms in $\relenv$ are intertwined. As for set
interpretations, we know from~\cite{jp19} that, for every $\Gamma;
\ol{\alpha} \vdash F$, $\relsem{\Gamma; \ol{\alpha} \vdash F}$ is
actually in $[\rel^k,\rel]$ where $k = |\ol \alpha|$.
%In fact, we already know from~\cite{jp19} that, for every $\Gamma;
%\ol{\alpha} \vdash \tau : \F$, $\relsem{\Gamma; \ol{\alpha} \vdash
%  \tau}$ is actually functorial in $\ol{\alpha}$ and
%$\omega$-cocontinuous.
We first define the actions of each of these functors on morphisms
between environments in Definition~\ref{def:rel-sem-funcs}, and then
argue that the functors given by Definitions~\ref{def:rel-sem}
and~\ref{def:rel-sem-funcs} are well-defined and have the required
properties. To do this, we extend $T_H$ to a {\em functor} from
$\relenv$ to $[[\rel^k,\rel],[\rel^k,\rel]]$. Its action on an object
$\rho \in \relenv$ is given by the higher-order functor $T_{H,\rho}$
whose actions on objects and morphisms are given in
Definition~\ref{def:rel-sem-funcs}. Its action on a morphism $f : \rho
\to \rho'$ is the higher-order natural transformation $T_{H,f} :
T_{H,\rho} \to T_{H,\rho'}$ whose action on any $F : [\rel^k,\rel]$ is
the natural transformation $T_{H,f}\, F : T_{H,\rho}\, F \to
T_{H,\rho'}\, F$ whose component at $\ol{R}$ is $(T_{H,f}\,
F)_{\ol{R}} = \relsem{\Gamma; \ol\gamma,\phi,\ol{\alpha} \vdash
  H}f[\phi := \id_F][\ol{\alpha := \id_R}]$. The next definition uses
$T_H$ to define the actions of functors interpreting types on
morphisms between relation environments.

@@

Straightforward calculation shows that terms of $\Nat$ type behave as

@@

If\, $\Gamma; \ol\beta \,|\, \Delta \vdash s : \Nat^{\overline{\alpha},
  \overline{\gamma}} F\,G$ and $\overline{\Gamma; \ol\beta \,|\,
  \Delta \vdash t : \Nat^{\overline{\gamma}} K\, H}$, then

@@

  \hspace*{0.135in}
\setsem{\Gamma; \ol\beta \,|\, \Delta
  \vdash L_{\overline{\gamma}} z. 
  (\map_G^{\overline{K}, \overline{H}} \,\overline{t})_{\ol\gamma}
  (s_{\overline{K}, \overline{\gamma}} z)
  : \Nat^{\overline{\gamma}} F[\overline{\alpha := K}]\,

@@

= \setsem{ \Gamma; \ol\beta \,|\, \Delta \vdash
  L_{\overline{\gamma}} z. s_{\overline{H}, \overline{\gamma}}
  ((\map_F^{\overline{K}, \overline{H}} \,
  \overline{t})_{\ol\gamma} z) : \Nat^{\overline{\gamma}} F[\overline{\alpha :=

@@

When $s = \mathsf{in}_H$, and
$\overline{\Gamma; \ol\delta \,|\, t :
 \Nat^{\overline{\gamma}} F\, G}$,
and $\xi = \Nat^{\overline{\gamma}} H[\phi :=

@@

\overline{\alpha}. H)\overline{G}$, Theorem~\ref{eq:ft-from-nat}

@@

\hspace*{0.135in}\setsem{\Gamma; \ol\delta \,|\, \Delta \vdash
L_{\overline{\gamma}} z. (\map_{(\mu \phi. \lambda

@@

    \overline{G}}\, \overline{t})_{\ol\gamma}
  (\tin_{H})_{\overline{F},

@@

= \setsem{\Gamma; \ol\delta \,|\, \Delta
\vdash L_{\overline{\gamma}} z. (\tin_H)_{\overline{G}, \overline{\gamma}}
  ((\map_{H[\phi := (\mu \phi. \lambda \overline{\alpha}. H)

@@

    \overline{G}}\, \overline{t})_{\ol\gamma} z)

@@

  &\setsem{\emptyset; \emptyset~|~ y : \Nat^\emptyset F \, G
    \vdash L_{{\emptyset}}
    z. (\map_{\mathit{Bush}\,\alpha}^{{F},
      {G}}\, y)_{\emptyset} (\mathit{reverseBush}_{F}\, z) : \Nat^{{\emptyset}}

@@

  &\setsem{\emptyset; \emptyset~|~ y : \Nat^\emptyset F\, G
    \vdash L_{\emptyset} z. \mathit{reverseBush}_{G}
    ((\map_{\mathit{Bush} \, \alpha}^{F, G}\,
           {y})_{\emptyset} z) : \Nat^{\emptyset} (\mathit{Bush} \, F ) \,

@@

  &\setsem{\emptyset; \emptyset~|~y : \Nat^\gamma F \, G 
  \vdash L_{{\gamma}} z.
    (\map_{\mathit{Tree}\,\alpha \, \gamma}^{F,
      G}\, y)_{\gamma} (\mathit{reverseTree}_{F, \gamma}\, z) : \Nat^{{\gamma}}

@@

    \gamma) }\\ =\; &\setsem{\emptyset; \emptyset~|~y :
    \Nat^\gamma F \, G \vdash L_{\gamma}
    z. \mathit{reverseTree}_{G, \gamma}\,
    ((\map_{\mathit{Tree} \, \alpha\, \gamma}^{{F},
      {G}}\, y)_{\gamma} z) : \Nat^{\gamma} (\mathit{Tree} \,

@@

Define, for $\Gamma; \ol{\alpha} \vdash F$, the term $\id_F$ to be
$\Gamma;\emptyset~|~\emptyset \vdash L_{\ol{\alpha}}x.x :
\Nat^{\ol{\alpha}} F\,F$ and, for terms $\Gamma; \emptyset \,|\,
\Delta \vdash t: \Nat^{\overline{\alpha}} F\,G$ and $\Gamma; \ol\beta
\,|\, \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H$, the {\em
  composition} $s \circ t$ of $t$ and $s$ to be $\Gamma; \ol\beta
\,|\, \Delta \vdash L_{\overline{\alpha}}
x. s_{\overline{\alpha}}(t_{\overline{\alpha}}x):
\Nat^{\overline{\alpha}} F\,H$. Then $\setsem{\Gamma; \emptyset \,|\,
  \emptyset \vdash \id_{F} : \Nat^{\ol{\alpha}} F\,F} \rho\, \ast =
\id_{\lambda \ol{A}. \setsem{\Gamma; \ol{\alpha} \vdash F} \rho
  [\ol{\alpha := A}]}$ for any set environment $\rho$, and
$\setsem{\Gamma; \ol\beta \,|\, \Delta \vdash s \circ t:
  \Nat^{\overline{\alpha}} F\,H} = \setsem{\Gamma; \ol\beta \,|\,
  \Delta \vdash s: \Nat^{\overline{\alpha}} G\,H} \circ
\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t:
  \Nat^{\overline{\alpha}} F\,G}$.  We have the expected relationships
between the interpetations of $\map$, $\tin$, and $\fold$:

@@

@@

\[\hspace*{-3.6in} = \setsem{\Gamma; \emptyset \,|\, \emptyset \vdash \mathit{id}_{(\mu
    \phi. \lambda \overline{\alpha}. H)\overline{\beta}} : \xi}\]

\vspace*{0.15in}

@@

\[\hspace*{-3.32in} =\setsem{ \Gamma; \emptyset \,|\, \emptyset \vdash
  \mathit{id}_{H[\phi := (\mu \phi. \lambda
      \overline{\alpha}. H)\overline{\beta}]} : \xi}\]

@@

%test
